Report on the Fourth Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE4)

### 2017-04-09 WSSSPE4_report/WSSSPE4_report.tex

\item[] {\bf Myth \#3: With massive migration of software systems to cloud-based environments we do not need to invest in a ``greener cloud'' anymore}. The use of cloud-based services promises large-scale optimizations in terms of the resources used by our software applications and related investments~\cite{green-cloud-sw}. Unfortunately, cloud provisioning heavily relies on over-capacity rather than intelligent governance, and so far has invested very little in innovation for sustainability. We need to build smarter software that both manages resources more efficiently in the cloud, and better understands the changes in the software operational context (including user requirements) and dynamically adapt to address sustainability intents~\cite{Procaccianti2016-2} (both in the cloud and on customer's premises.)\\

@@

Going into WSSSPE4, the WSSSPE organization had not had a formal mission or vision statement.
The organizers developed a strawhorse, which was presented to the participants early in the meeting.
The presentation included guidelines as well as examples from other similar communities.
The guidelines were that in general, an organization's mission should stand the test of time and state what is wrong and how the organization is going to fix it; Its vision should imagine what the world look like if the organization is successful; and from which, focus areas could be use to establish the scope of the organization along with its goals.

@@

{\bf Mission}
WSSSPE is an international community-driven organization that promotes sustainable research software by addressing challenges related to the full lifecycle of research software through shared learning and community action.

{\bf Vision}
We envision a world where research software is accessible, robust, sustained, and recognized as a scholarly research product critical to the advancement of knowledge, learning, and discovery.

{\bf Focus areas}
WSSSPE promotes sustainable research software by positively impacting:
\begin{itemize}
\item {\bf Principles and Best Practices}. Promoting best practices in sustainable software
\item {\bf Careers}. Developing and supporting career paths in research software development and engineering
\item {\bf Learning}. Engaging in activities to promote peer learning and interaction
\item {\bf Credit}. Ensuring recognition of research software as an intellectual contribution equal to other research products
\end{itemize}

@@

\item \textbf{Olivier Philippe, Simon Hettrick, and Neil Chue Hong:
\emph{Experience Paper: Preliminary analysis of a survey of UK Research
Software Engineers}}~\cite{Philippe:2016ws}.
This paper presents results from a survey conducted on the new role of research
software engineers in academia. The survey provides demographic information
about the education, field, gender, job satisfaction and career plans of the people
in this community. The outcome is that members of the community  were found to be
highly educated, derive mainly from the hard sciences, and are predominantly male.
Respondents reported satisfaction in their jobs, but indicated that career progression
is both difficult and opaque. The authors urge a continued discussion about the experience
of research software engineers and recommend further investigation into this
important community.

@@

\item \textbf{Neil Chue Hong: \emph{Position Paper: Why do we need to compare
research software, and how should we do it?}}~\cite{ChueHong:2016ws}.
This position paper sets out the reasons for why different stakeholders,
from users to developers to funders, might wish to undertake the difficult task
to compare research software. The author describes a proposed framework for doing
so (based around measures of accessibility, usability, maintainability and
portability) which takes into account the possibility of variation between
different communities about how they prioritize different aspects of research software.

@@

    \item \textbf{Daniel S.~Katz, Kyle E.~Niemeyer\textsuperscript{\textasteriskcentered},
    Arfon M.~Smith, and the FORCE11 Software Citation Working Group:
    \emph{Software Citation: Process, Principles, and Implementation}}~\cite{Katz:2016ws}.
    This talk described the timeline of activities of the FORCE11 Software Citation
    Working Group and the principles of software citation established by the group.
    Niemeyer, one of the working group cochairs, also discussed the paper published
    describing the principles~\cite{Smith:2016sc}, and the remaining steps of the
    working group before sunsetting it and spinning up a new Software Citation
    Implementation Working Group.

@@

To investigate best practices for scientific software we brought together a panel of five experts with different perspectives on ``Best Practices for Scientific Software.''
These were: Alice Allen, Editor, Astrophysics Source Code Library, brought an understanding of the difficulties of organizing a community and curating their software.
Mike Croucher from the University of Sheffield and Rob Haines from the University of Manchester who are both Research Software Engineers with decades of experience of writing code for researchers.
Patricia Lago from Vrije Universiteit Amsterdam presented a keynote at the workshop (\S\ref{sec:keynote}) and brought a fresh perspective on software sustainability from the point of view of its impact on society and business.
Finally, Karthik Ram from the Berkeley Institute for Data Science brought his perspective on the practicalities of using scientific software to conduct his research as a data scientist.

The panel opened with a question to the floor ``Who in the audience still codes in anger?'' which caused some confusion as, it turns out, that this is a turn of phrase that has not crossed the Atlantic.
After translating this question into the more open ``Who in the audience still writes code as part of their day-to-day work?'' we were able to confirm the suspicion that the WSSSPE audience is heavily involved writing, not just using, code.
Hence, the panel were going to have to deal with questions from a highly experienced audience.

@@

The groups worked in Google docs, and eventually created Google slides to show their work back to the larger groups. The docs and slides were stored in a common Google Drive folder\footnote{\url{https://drive.google.com/drive/u/0/folders/0BzIpChHwv-q5X3lRaWhYMXJZaUE}}. In each group's discussion, a number of Specific, Measurable, Achievable, Relevant and Timely (SMART) steps were created, and ideally, at least the first one (to set up some communication mechanism) was completed.  In many cases, this was done through Slack (see \S\ref{sec:slack}.)

>>>>

\item[] {\bf Myth \#3: With massive migration of software systems to cloud-based environments, we do not need to invest in a ``greener cloud'' anymore}. The use of cloud-based services promises large-scale optimizations in terms of the resources used by our software applications and related investments~\cite{green-cloud-sw}. Unfortunately, cloud provisioning heavily relies on over-capacity rather than intelligent governance, and so far has invested very little in innovation for sustainability. We need to build smarter software that both manages resources more efficiently in the cloud, better understands the changes in the software operational context (including user requirements), and dynamically adapts to address sustainability intents~\cite{Procaccianti2016-2} (both in the cloud and on customer's premises.)\\

@@

Going into WSSSPE4, the WSSSPE organization had not had a formal mission or vision statement.
The organizers developed a strawhorse, which was presented to the participants early in the meeting.
The presentation included guidelines as well as examples from other similar communities.
The guidelines were that in general, an organization's mission should stand the test of time and state what is wrong and how the organization is going to fix it; its vision should imagine what the world look like if the organization is successful; and from which, focus areas could be use to establish the scope of the organization along with its goals.

@@

{\bf Mission.}
WSSSPE is an international community-driven organization that promotes sustainable research software by addressing challenges related to the full lifecycle of research software through shared learning and community action.

{\bf Vision.}
We envision a world where research software is accessible, robust, sustained, and recognized as a scholarly research product critical to the advancement of knowledge, learning, and discovery.

{\bf Focus areas.}
WSSSPE promotes sustainable research software by positively impacting:
\begin{itemize}
\item {\bf Principles and Best Practices}. Promoting best practices in sustainable software
\item {\bf Careers}. Developing and supporting career paths in research software development and engineering
\item {\bf Learning}. Engaging in activities to promote peer learning and interaction
\item {\bf Credit}. Ensuring recognition of research software as an intellectual contribution equal to other research products
\end{itemize}

@@

\item \textbf{Olivier Philippe, Simon Hettrick, and Neil Chue Hong:
\emph{Experience Paper: Preliminary analysis of a survey of UK Research
Software Engineers}}~\cite{Philippe:2016ws}.
This paper presents results from a survey conducted on the new role of research
software engineers in academia. The survey provides demographic information
about the education, field, gender, job satisfaction, and career plans of the people
in this community. The outcome is that members of the community  were found to be
highly educated, derive mainly from the hard sciences, and are predominantly male.
Respondents reported satisfaction in their jobs, but indicated that career progression
is both difficult and opaque. The authors urge a continued discussion about the experience
of research software engineers and recommend further investigation into this
important community.

@@

\item \textbf{Neil Chue Hong: \emph{Position Paper: Why do we need to compare
research software, and how should we do it?}}~\cite{ChueHong:2016ws}.
This position paper sets out the reasons for why different stakeholders,
from users to developers to funders, might wish to undertake the difficult task
to compare research software. The author describes a proposed framework for doing
so (based around measures of accessibility, usability, maintainability, and
portability) which takes into account the possibility of variation between
different communities about how they prioritize different aspects of research software.

@@

    \item \textbf{Daniel S.~Katz, Kyle E.~Niemeyer\textsuperscript{\textasteriskcentered},
    Arfon M.~Smith, and the FORCE11 Software Citation Working Group:
    \emph{Software Citation: Process, Principles, and Implementation}}~\cite{Katz:2016ws}.
    This talk described the timeline of activities of the FORCE11 Software Citation
    Working Group and the principles of software citation established by the group.
    Niemeyer, one of the working group co-chairs, also discussed the paper published
    describing the principles~\cite{Smith:2016sc}, and the remaining steps of the
    working group before sunsetting it and spinning up a new Software Citation
    Implementation Working Group.

@@

To investigate best practices for scientific software we brought together a panel of five experts with different perspectives on ``Best Practices for Scientific Software.''
These were: Alice Allen, Editor, Astrophysics Source Code Library, brought an understanding of the difficulties of organizing a community and curating their software;
Mike Croucher from the University of Sheffield and Rob Haines from the University of Manchester who are both Research Software Engineers with decades of experience of writing code for researchers;
Patricia Lago from Vrije Universiteit Amsterdam presented a keynote at the workshop (\S\ref{sec:keynote}) and brought a fresh perspective on software sustainability from the point of view of its impact on society and business; and
Karthik Ram from the Berkeley Institute for Data Science brought his perspective on the practicalities of using scientific software to conduct his research as a data scientist.

The panel opened with a question to the floor ``Who in the audience still codes in anger?'' which caused some confusion as, it turns out, that this is a turn of phrase that has not crossed the Atlantic.
After translating this question into the more open, ``Who in the audience still writes code as part of their day-to-day work?'' we were able to confirm the suspicion that the WSSSPE audience is heavily involved writing, not just using, code.
Hence, the panel were going to have to deal with questions from a highly experienced audience.

@@

The groups worked in Google docs, and eventually created Google slides to show their work back to the larger groups. The docs and slides were stored in a common Google Drive folder\footnote{\url{https://drive.google.com/drive/u/0/folders/0BzIpChHwv-q5X3lRaWhYMXJZaUE}}. In each group's discussion, a number of Specific, Measurable, Achievable, Relevant, and Timely (SMART) steps were created, and ideally, at least the first one (to set up some communication mechanism) was completed.  In many cases, this was done through Slack (see \S\ref{sec:slack}.)

### 2017-04-09 WSSSPE4_report/wg_codemeta.tex

Research software is often not shared; that that is shared may not have much metadata associated with it, and that which does exist often does not travel further than the website on which the software resides. CodeMeta wants to incentivize software developers to release their software, encourage the development of metadata for it, enable credit assignment and citation of research software and increase its discoverability, more easily track dependencies, and enable reuse of software metadata, all goals that WSSSPE attendees have great interest in supporting.

>>>>

Research software is often not shared; that which is shared may not have much metadata associated with it, and that which does exist often does not travel further than the website on which the software resides. CodeMeta wants to incentivize software developers to release their software, encourage the development of metadata for it, enable credit assignment and citation of research software, increase its discoverability, more easily track dependencies, and enable reuse of software metadata, all goals that WSSSPE attendees have great interest in supporting.
